---
title: "Reading, Writing and Exploring Data"
output:
  beamer_presentation: default
  ioslides_presentation: default
---

```{r setup, echo=FALSE}
knitr::opts_knit$set(root.dir = "C:/RFiles")
```

##Exercise files

Please download all exercise files for this exercise at http://georgejmount.com/hillsdaler.


##Loading our libraries

It's a good idea to call the libraries you're going to use in your script at the beginning of the program. 

We will be using two libraries,`tidyverse` and `readxl`. Load them now:

```{r eval=FALSE}
library(tidyverse)
library(readxl)
```

If you see an error message, check your packages are installed: **`install.packages()`**.

#DRILL

1. Install and load the `caret` package to your machine.

---

## Reading in Files from R

While R has its own file extension for storing data, it's more common to reading and writing to outside file types. We will cover how to read and write to `.csv`, `.txt` and `.xlsx` files.

we will be importing in the famous `iris` dataset which is popular in data science training. 

The data set consists of 50 samples from each of three species of Iris (Iris setosa, Iris virginica and Iris versicolor). Four features were measured from each sample: the length and the width of the sepals and petals, in centimetres.

## Working Directories in R

An active R session always has a working directory associated with it. This is where, by default, all files are read and write to. 

Check the location of your current directory with the `getwd` function:

```{r eval=FALSE}
getwd()
```

##`getwd`

You can set a working directory with `setwd()`. 

File paths in R are always enclosed in double quotation marks, and R uses forward slashes, not backslashes, when specifying folder locations. This will take some getting used to, especially for Windows users!

```{r eval=FALSE}
setwd("C:/RFiles/")
```


## Relative and absolute file paths

There are two ways to refer to outside files in R: using relative or absolute file paths.

Which to use? It all depends on whether you have set a **working directory** with `setwd()` and whether the file you want to read/write to is in that folder.

For example, let's say I have the iris dataset saved in the folder `C:/RFiles/` **and** I have set the working directory to the same folder.

In this case I will use the **relative** file path...

```{r eval=FALSE, COLLAPSE=TRUE}
setwd("C:/RFiles")
iris<-read_csv("iris.csv")
```

----

Let's say that instead I had set my working directory to another folder, `"C:/Iris"`.

In this case a relative reference will not work. 



```{r error=TRUE, COLLAPSE=TRUE, eval=FALSE}
setwd("C:/Iris")
iris<-read_csv("iris.csv")
```


Instead, we need to use the *absolute* file path because the file is not in the working directory.


```{r error=TRUE, echo=FALSE, eval=FALSE, collapse=TRUE}
#We will use the absolute file path becuase 
#the file is not in the current working directory
iris<-read_csv("C:/RFiles/iris.csv")
```

----

For the remainder of this unit, I will save all files to `C:/RFiles` which will also be my working directory:

```{r error=TRUE, eval=FALSE, collapse=TRUE}
setwd("C:/RFiles")
```


##Reading and Writing to CSV Files

To *read* files we will use `read_csv()`

*Make sure to assign this function to an object!* If not, you will simply have read the file in R without any way to call it!

```{r error=TRUE, echo=FALSE, eval=FALSE, collapse=TRUE}
#assign the iris.csv dataset to iris variable
iris<-read_csv("iris.csv")
```


To *write* files we will use (you guessed it!) `write_csv()`. 

This function will take one extra argument in that we first type the object where the dataset is pointed to and then write to the file we want.

----

However, we will not need to assign it to an object, since we are exporting it anyway.

```{r error=TRUE, eval=FALSE, COLLAPSE=TRUE}
#write the iris dataframe to iriswrite.csv
write_csv(iris,"iriswrite.csv")
```

This file will now appear in our *working directory* (`C:/RFiles`).

##Reading and writing to .txt files

This will be very similar to the `read_csv` and `write_csv` functions: We use `read_tsv` and `write_tsv`.

```{r error=TRUE, eval=FALSE, collapse=TRUE}
#assign the iris.txt dataset to iris variable
iris<-read_tsv("iris.txt")
```

```{r error=TRUE, eval=FALSE, collapse=TRUE}
#write the iris dataset to Iriswrite.txt
iris<-write_tsv("iriswrite.txt")
```

##Reading and Writing Excel Files

We will now read in the Iris.xlsx file using the `read_excel` function:

```{r error=TRUE, eval=FALSE, collapse=TRUE}
#read file to excel
iris<-read_excel("iris.xlsx")
```

To write files to Excel, we will... **write to csv**.

Unfortunately, the `readxl` package only reads files from Excel, as the name suggests. There are packages to write to Excel, but I find that writing to CSV and then opening in Excel is the easiest way. 


```{r error=TRUE, eval=FALSE, collapse=TRUE}
#write to csv and then open in Excel
write_csv(iris,"C:/RFiles/iriswrite2.csv")
```


##A sneak peek at your data

If you are used to exploring your data visually, as you might when scrolling down a spreadsheet, you can use `view()`.

```{r, eval=FALSE,collapse=TRUE}
#open up spreadsheet-type environment
view(iris)
```

Because you're often dealing with such huge datasets, it may not be practical to try viewing the entire table. 

##`head` and `tail`

Instead, use `head` and `tail` to print the first and last X rows, respectively.


```{r collapse=TRUE}
#view first 3 observations
head(iris,3)
```

```{r collapse=TRUE}
#view last 3 observations
tail(iris,3)
```

##A brief detour of tibbles

You may have noticed your output for the previous functions began with: `A tibble: 3 x 5`.

What exactly is a tibble?

Put simply, it is a 'modern' data frame used by the libraries of the `tidyverse`. They are nearly identical in concept - tibbles are just programmed to work better for `tidyverse` libraries.

Back to our scheduled programming (heh, heh)...

##`str`

The `str()` function gives us an idea of the structure and type of the object. 

```{r collapse=TRUE}
str(iris)
```

##`names`

If you just want a list of the names in the dataset, you can run the `names()` function.

```{r collapse=TRUE}
names(iris)
```

##`summary`

Finally, you can get summary statistics of your variables with the `summary()` function:

```{r collapse=TRUE}
summary(iris)
```

#Questions?